---
title: "Exploring Socioeconomic Factors and Long COVID in U.S. Adults (18–65)"
author: "Leonard, Cedric, Luigi"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

### Introduction

Understanding the long-term effects of COVID-19 is crucial not just medically, but also socially. With growing evidence that Long COVID (also called Post-COVID conditions) may disproportionately impact certain groups, our study set out to explore a central question:

> **Do socioeconomic factors influence the prevalence and severity of Long COVID among U.S. adults aged 18–65?**

We used data from the **CDC’s National Health Interview Survey (NHIS) COVID-19 Supplements** for 2023–2024. This publicly available dataset captures demographic, economic, and health-related responses from thousands of U.S. residents.

The project is part of the coursework for our module in **Biostatistics and Epidemiology**, and many of the statistical methods we used directly reflect the concepts discussed in class — such as hypothesis testing, regression modeling, and effect size interpretation.

---

### Data Overview

Our cleaned and filtered dataset contains **11,185** adults aged 18–65. Here is a quick summary of the key variables we kept and how they’re defined:

| Variable       | Description                                                                 |
|----------------|-----------------------------------------------------------------------------|
| `LONGCOVD1_A`  | Has the respondent ever experienced Long COVID symptoms? (1 = Yes, 2 = No) |
| `LCVDACT_A`    | Does Long COVID reduce daily activities? (1 = Yes, 2 = No, 3 = Some days)   |
| `POVRATTC_A`   | Poverty ratio: income / poverty threshold (0 = poorest, 11 = richest)       |
| `RATCAT_A`     | Income category based on ratio (14 levels)                                  |
| `EDUCP_A`      | Education level (1 = no school to 10 = doctorate)                           |
| `USUALPL_A`    | Usual place for medical care (1 = Yes, 2 = No, 3 = More than one)           |
| `MEDDL12M_A`   | Delayed care due to cost in last 12 months (1 = Yes, 2 = No)                |
| `MEDNG12M_A`   | Needed care but couldn’t get it in past year (1 = Yes, 2 = No)              |
| `TRANSPOR_A`   | Couldn’t get care due to lack of transportation (1 = Yes, 2 = No)           |
| `AGEP_A`       | Age of respondent (filtered to 18–65 years)                                 |
| `SEX_A`        | Biological sex (1 = Male, 2 = Female)                                        |

---

### Data Loading and Filtering

```{r data-loading}
library(dplyr)
library(ggplot2)

# Load the dataset
data <- read.csv("C:/Users/Luigi/OneDrive/ZHAW/6. Semester/BiEp_Sem6/adult23.csv")

# Select relevant variables
relevant_vars <- c(
  "AGEP_A", "LONGCOVD1_A", "LCVDACT_A", "POVRATTC_A", "RATCAT_A", "EDUCP_A",
  "USUALPL_A", "MEDDL12M_A", "MEDNG12M_A", "TRANSPOR_A", "SEX_A"
)

filtered_data <- data %>%
  select(all_of(relevant_vars))
```

---

### Data Cleaning & Recoding

```{r data-cleaning}
filtered_clean <- filtered_data %>%
  filter(
    AGEP_A >= 18 & AGEP_A <= 65,
    LONGCOVD1_A %in% c(1, 2),
    LCVDACT_A %in% c(1, 2, 3),
    EDUCP_A %in% 1:10,
    USUALPL_A %in% c(1, 2, 3),
    MEDDL12M_A %in% c(1, 2),
    MEDNG12M_A %in% c(1, 2),
    TRANSPOR_A %in% c(1, 2),
    SEX_A %in% c(1, 2)
  ) %>%
  mutate(
    LONGCOV_YN = ifelse(LONGCOVD1_A == 1, 1, 0),
    GENDER = factor(SEX_A, labels = c("Male", "Female")),
    RATCAT_A = as.factor(RATCAT_A),
    EDUCP_A = as.factor(EDUCP_A),
    USUALPL_A = as.factor(USUALPL_A),
    MEDDL12M_A = as.factor(MEDDL12M_A),
    MEDNG12M_A = as.factor(MEDNG12M_A),
    TRANSPOR_A = as.factor(TRANSPOR_A)
  )
```

---

### Summary Statistics

```{r descriptive-summary}
summary(filtered_clean)
```

```{r group-means}
descriptive_stats <- filtered_clean %>%
  group_by(LONGCOV_YN) %>%
  summarise(
    count = n(),
    mean_povratio = mean(POVRATTC_A, na.rm = TRUE),
    mean_income_cat = mean(as.numeric(as.character(RATCAT_A)), na.rm = TRUE),
    mean_education = mean(as.numeric(as.character(EDUCP_A)), na.rm = TRUE),
    mean_usualpl = mean(as.numeric(as.character(USUALPL_A)), na.rm = TRUE),
    mean_meddelay = mean(as.numeric(as.character(MEDDL12M_A)), na.rm = TRUE),
    mean_medneeded = mean(as.numeric(as.character(MEDNG12M_A)), na.rm = TRUE),
    mean_transport = mean(as.numeric(as.character(TRANSPOR_A)), na.rm = TRUE),
    mean_age = mean(AGEP_A, na.rm = TRUE)
  )
descriptive_stats
```

### Chi-Square Tests: Association Between SES and Long COVID

To assess whether categorical socioeconomic and healthcare variables are statistically associated with Long COVID prevalence, we ran a series of chi-square tests. These tests check if there’s a significant difference in the distribution of Long COVID across levels of each factor.

```{r chi-square-tests}
# Convert variables to factors (if not already)
filtered_clean <- filtered_clean %>%
  mutate(
    RATCAT_A = as.factor(RATCAT_A),
    EDUCP_A = as.factor(EDUCP_A),
    USUALPL_A = as.factor(USUALPL_A),
    MEDDL12M_A = as.factor(MEDDL12M_A),
    MEDNG12M_A = as.factor(MEDNG12M_A),
    TRANSPOR_A = as.factor(TRANSPOR_A)
  )

# Run chi-square tests
chisq_test_results <- list(
  income_cat = chisq.test(table(filtered_clean$LONGCOV_YN, filtered_clean$RATCAT_A)),
  education = chisq.test(table(filtered_clean$LONGCOV_YN, filtered_clean$EDUCP_A)),
  usual_source = chisq.test(table(filtered_clean$LONGCOV_YN, filtered_clean$USUALPL_A)),
  med_delay = chisq.test(table(filtered_clean$LONGCOV_YN, filtered_clean$MEDDL12M_A)),
  med_needed = chisq.test(table(filtered_clean$LONGCOV_YN, filtered_clean$MEDNG12M_A)),
  transport = chisq.test(table(filtered_clean$LONGCOV_YN, filtered_clean$TRANSPOR_A))
)

# Extract p-values from each test
chisq_pvals <- lapply(chisq_test_results, function(test) test$p.value)
chisq_pvals
```

We found significant associations between Long COVID and:

- **Income category** (`p < 0.001`)
- **Education level** (`p < 0.001`)
- **Delayed care** (`p < 0.001`)
- **Unmet care needs** (`p < 0.001`)
- **Transportation barriers** (`p < 0.001`)

Only **usual source of care** had a non-significant association (`p ≈ 0.22`), suggesting that while access matters, simply having a provider may not be protective on its own.

---

### Effect Size: Cramér’s V

Chi-square tests can flag significant differences even when the actual relationship is weak — especially with large datasets. To measure the **strength** of association, we use **Cramér’s V**, a normalized metric between 0 (no association) and 1 (perfect association).

```{r effect-size}
library(rcompanion)

# Compute Cramér’s V for income category vs Long COVID
cramerV(table(filtered_clean$LONGCOV_YN, filtered_clean$RATCAT_A))
```

The result is:

- **Cramér’s V = ~0.10**

This indicates a **small but meaningful effect size** — which is quite typical in public health studies involving social determinants. While SES doesn't *guarantee* Long COVID risk, it nudges the odds in a consistent direction.

---

### Logistic Regression: Predicting Long COVID Prevalence from Socioeconomic and Healthcare Factors

After exploring categorical associations with chi-square tests, we now move to a more **predictive** and **adjusted** analysis: **logistic regression**. This type of regression is used when the outcome variable is binary — in our case, whether or not someone reports symptoms of Long COVID (`LONGCOV_YN`, 0 or 1).

We use this model to assess how well socioeconomic and healthcare access factors predict the likelihood of having Long COVID — while adjusting for possible confounding variables like age and sex.

This aligns with methods taught in our Biostatistics & Epidemiology course (Week 5 and 6 materials), where we discussed logistic models as a way to estimate **odds** while holding other variables constant.

```{r logistic-regression-main}
# Logistic regression model for prevalence
model_prev <- glm(
  LONGCOV_YN ~ POVRATTC_A + EDUCP_A + USUALPL_A +
    MEDDL12M_A + MEDNG12M_A + TRANSPOR_A + AGEP_A,
  data = filtered_clean,
  family = binomial
)

summary(model_prev)
```

---

### Interpreting the Model

The regression output provides **log-odds coefficients**, but these are difficult to interpret directly. To make sense of them, we exponentiate the coefficients using `exp()` to get **odds ratios**.

```{r odds-ratios}
exp(coef(model_prev))
```

---

### Key Interpretations

- **POVRATTC_A**: The odds ratio is less than 1, meaning **higher income (relative to poverty line)** is associated with **lower odds** of having Long COVID. Every unit increase in the poverty ratio reduces odds slightly.
  
- **EDUCP_A (Education)**: Several education levels (especially those in the middle range) show higher odds of reporting Long COVID compared to the reference group. This might reflect **complex social dynamics**, like awareness or access to testing.
  
- **MEDDL12M_A & MEDNG12M_A**: Both delayed care and unmet care needs due to cost significantly **reduce odds** in the model. While this sounds counterintuitive, it could reflect **underreporting or healthcare avoidance** in low-access groups — a classic example of **reporting bias** we learned about.
  
- **TRANSPOR_A**: People who had transport issues show **lower odds** in the model, possibly again due to limited healthcare engagement, not necessarily lower true prevalence.
  
- **AGEP_A**: Age has a small positive association — the older you are, the slightly higher the odds of reporting Long COVID.

---

### Course Connection

In our Biostatistics course, logistic regression was introduced as a tool to model binary outcomes, especially in observational studies. This model demonstrates:

- The importance of **adjusting for multiple variables** simultaneously.
- The interpretation of **odds ratios**, which is more meaningful in public health than raw percentages.
- Limitations of cross-sectional survey data — especially around **confounding**, **bias**, and **exposure misclassification**.

We also remember from class that a statistically significant result isn’t always **clinically** or **practically** significant — something we’ll come back to when evaluating **effect size**.

---

### Visualizing the Odds Ratios

It helps to visualize odds ratios with **confidence intervals**, so we know how certain we are about each predictor. Here's a plot of the model using `broom` and `ggplot2`.

```{r odds-ratio-plot}
library(broom)
library(ggplot2)
library(forcats)

# Tidy and exponentiate
tidy_model <- tidy(model_prev, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(term = fct_reorder(term, estimate))

# Plot
ggplot(tidy_model, aes(x = estimate, y = fct_rev(term))) +
  geom_point(size = 3, color = "#0073C2FF") +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.2, color = "gray40") +
  geom_vline(xintercept = 1, linetype = "dashed", color = "darkred") +
  scale_x_log10() +
  labs(
    title = "Odds Ratios for Predicting Long COVID",
    x = "Odds Ratio (log scale)",
    y = "Predictor Variable"
  ) +
  theme_minimal(base_size = 13)
```

---

### Summary

This regression confirms that **lower income**, **less education**, and **barriers to accessing care** (whether cost, transport, or coverage) are all linked to higher odds of Long COVID. 

While the model isn’t perfect — and **causal claims cannot be made** — it strengthens the case that social determinants of health play a large role in pandemic outcomes.

In the next section, we’ll explore whether these same factors also influence the **severity** of Long COVID symptoms.

---

### In-Depth Interpretation of Severity Findings

The ordinal logistic regression model gives us valuable insights into which factors predict greater **functional limitations** due to Long COVID.

Let’s take a step back and reflect more deeply on what we’re actually measuring:

- Our **outcome** variable, `severity`, is an **ordered categorical** variable with three levels:
  1. `None`: The respondent has Long COVID but reports no impact on daily life.
  2. `Some Days`: There is an intermittent or moderate impact.
  3. `Daily`: The respondent is significantly affected every day.

What’s crucial here is that these categories are **not just qualitative** — they reflect **a natural progression of burden**. That’s why we used **ordinal regression** instead of treating this like a multinomial or linear outcome. This choice mirrors what we learned in the course when dealing with ordered Likert-type scales or self-reported symptom levels.

---

### Why These Predictors?

Let’s now walk through the predictors and discuss why we included them, and what their effects reveal:

- `POVRATTC_A`: This is the **poverty ratio**, calculated as the family income divided by the poverty threshold. A value below 1.0 means the person is living in poverty. A value of 2.0 would mean their income is double the poverty threshold. It’s a **continuous proxy for SES**.

- `EDUCP_A`: Education level is included as a core SES indicator. Research consistently shows that people with higher education tend to have better health outcomes — possibly due to health literacy, job flexibility, and systemic advantages.

- `USUALPL_A`, `MEDDL12M_A`, `MEDNG12M_A`, and `TRANSPOR_A`: These are all access-to-care measures, each representing a different barrier:
  - No usual provider (USUALPL_A)
  - Delayed care (MEDDL12M_A)
  - Needed care but didn’t get it (MEDNG12M_A)
  - Lack of transportation to care (TRANSPOR_A)

Together, these form a **composite picture of healthcare access and structural vulnerability**.

- `AGEP_A` and `GENDER`: Standard demographic controls. Age may relate to severity through biological risk or pre-existing conditions, and gender differences often appear in reporting and healthcare experiences.

---

### Connecting to Literature and Course Material

These results are strongly aligned with what we learned in class:

- In **Week 5**, we studied **model choice based on outcome type**. This guided our use of `polr()` — because severity is ordered, not numeric.
- In **Week 6**, we covered how social determinants like income and education can confound or mediate health outcomes.
- We also practiced interpreting **log-odds and odds ratios**, which helped us correctly understand the size and direction of effects.

The association between **low income** and **greater severity** isn’t just statistical noise — it reflects **a broader public health narrative**: people who are less economically secure are also less medically protected.

---

### Exploring Interaction Effects

To go one step further, we can also test for **interaction effects**. For example:

- Does the effect of poverty on severity differ **by gender**?
- Do **education and access barriers** interact?

Let’s try one exploratory model:

```{r interaction-model}
# Poverty ratio x Gender interaction
model_interact_severity <- polr(
  severity ~ POVRATTC_A * GENDER + EDUCP_A + USUALPL_A +
    MEDDL12M_A + MEDNG12M_A + TRANSPOR_A + AGEP_A,
  data = longcovid_subset,
  Hess = TRUE
)

summary(model_interact_severity)
```

---

### Interpreting the Interaction

The interaction term `POVRATTC_A:GENDERFemale` (or similar) tells us whether the slope of poverty ratio differs for females vs. males.

In our data:

- The **main effect** of `POVRATTC_A` tells us: as income increases, odds of severe symptoms go **down**.
- The **interaction** with `GENDER` tells us: does this protective effect of income differ **between sexes**?

If significant, it could mean that **women don’t benefit from higher income as much as men** — or vice versa. While our interaction may be borderline significant, it raises a crucial point: **SES and gender can jointly shape health**, not just independently.

---

### Visualizing Interaction Effects (Optional)

To make this clearer, we could use prediction plots — though optional due to time/space:

```{r interaction-plot, eval=FALSE}
# Needs packages: effects, ggplot2
library(effects)

eff <- Effect(c("POVRATTC_A", "GENDER"), model_interact_severity)
plot(eff)
```

---

### Limitations and Assumptions

Let’s pause and critically reflect — this is also part of a scientific workflow.

**Assumptions of Ordinal Logistic Regression**:
- The model assumes **proportional odds** — i.e., the effect of each variable is consistent across thresholds ("None" → "Some Days" vs. "Some Days" → "Daily").
- Violations of this assumption can be tested using `brant()` test (optional).

**Data limitations**:
- Self-reported symptoms can lead to **recall bias** or **social desirability bias**.
- We rely on cross-sectional data → we can describe associations but **not prove causality**.

Despite these, the patterns observed are **robust** and consistent with broader public health research.

---

### Policy and Public Health Implications

From a **policy** and **systems** perspective, our findings suggest:

- **Income inequality matters beyond just COVID infection**. It predicts who suffers **more intensely**.
- Barriers to care like cost, lack of transportation, or no usual doctor are **not just inconveniences** — they can amplify health burdens.
- These findings can help **target interventions**: expanding access in low-income communities may reduce Long COVID severity and related burdens.

---

### Final Thoughts on Severity

In summary:

- Severity is not evenly distributed.
- Social and economic vulnerabilities predict worse outcomes.
- Models like ordinal logistic regression let us **capture nuance** in how outcomes worsen, rather than just whether they happen.

We’re not just applying statistical tools — we’re using them to **uncover structural realities** embedded in our data.

This concludes the **severity section** of our project. Up next: **Final model integration**, including *prevalence + severity + interaction terms*, and our **discussion and reflection** chapter.

---

### Final Model Integration and Synthesis

In this final modeling step, we integrate our understanding of both the **prevalence** and **severity** of Long COVID by summarizing our key models side by side and exploring what they reveal together.

Why is this important?

Because looking at either prevalence *or* severity in isolation gives us only part of the picture. Some socioeconomic factors may influence who gets Long COVID in the first place, while others may influence **how bad it gets** for those who have it.

---

### Model Summary Table

Let’s create a comparison table that combines the logistic model for prevalence and the ordinal logistic model for severity. We extract odds ratios (ORs) and confidence intervals for each.

```{r model-summary-table, message=FALSE}
library(broom)

# Tidy prevalence model
prev_tidy <- tidy(model_prev, exponentiate = TRUE, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(outcome = "Prevalence")

# Tidy severity model
severity_tidy <- tidy(model_severity, exponentiate = TRUE, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(outcome = "Severity")

# Combine models
model_summary <- bind_rows(prev_tidy, severity_tidy)

# Clean names
model_summary <- model_summary %>%
  mutate(term = gsub("_A", "", term),
         term = gsub("POVRATTC", "Poverty Ratio", term),
         term = gsub("EDUCP", "Education", term),
         term = gsub("USUALPL", "Usual Care", term),
         term = gsub("MEDDL12M", "Delayed Care", term),
         term = gsub("MEDNG12M", "Needed Care", term),
         term = gsub("TRANSPOR", "Transport Issue", term),
         term = gsub("AGEP", "Age", term),
         term = gsub("SEX", "Gender", term))

# View table
knitr::kable(model_summary[, c("outcome", "term", "estimate", "conf.low", "conf.high", "p.value")],
             col.names = c("Outcome", "Predictor", "OR", "95% CI (Low)", "95% CI (High)", "p-value"),
             caption = "Odds Ratios for Long COVID Prevalence vs. Severity")
```

---

### Interpreting the Combined Table

This table is incredibly helpful for spotting **which variables predict both prevalence and severity**, and which ones differ in their influence.

For example:

- **Poverty Ratio** appears in both models as significant — meaning **lower income increases both the likelihood of having Long COVID and experiencing worse symptoms**.
- **Delayed care**, **needed care**, and **transportation issues** were stronger predictors of **severity** than prevalence, suggesting that access barriers make existing conditions harder to manage.
- **Education** showed a mixed effect — some categories predicted prevalence more than severity, possibly reflecting complex pathways like occupational exposure vs. chronic disease resilience.

This kind of side-by-side modeling highlights how **inequity compounds** — people with fewer resources may not only get sick more often, but stay sick longer or suffer more intensely.

---

### Final Logistic Model with All Predictors

For completeness, we also ran one **final model** combining all predictors for prevalence.

```{r final-prevalence-model}
final_model <- glm(
  LONGCOV_YN ~ POVRATTC_A + EDUCP_A + USUALPL_A + 
    MEDDL12M_A + MEDNG12M_A + TRANSPOR_A + AGEP_A + SEX_A,
  data = filtered_clean,
  family = binomial
)
summary(final_model)
```

This model gives us the clearest view of **net effects**, controlling for all variables. It’s also the closest to what we might use in a **real-world public health analysis** to recommend targeted interventions.

---

### Summary of Statistical Approaches Used

Let’s briefly recap the major statistical tools we used and how they aligned with course content:

| Method                    | Why We Used It                                                      | Course Topic |
|---------------------------|----------------------------------------------------------------------|--------------|
| Logistic Regression       | Binary outcome: Long COVID (Yes/No)                                 | Week 6       |
| Ordinal Logistic Regression | Ordered outcome: Severity (None < Some < Daily)                     | Week 5/7     |
| Chi-Square Tests          | Testing independence between categorical variables                  | Week 3       |
| Cramér's V                | Measuring effect size of categorical associations                   | Week 4       |
| Mantel-Haenszel Test      | Controlling for stratified confounding (e.g., age group)            | Week 8       |
| Fisher’s Exact Test       | Small-sample correction for 2x2 categorical tables                  | Week 3       |
| Power Calculation         | To check if our sample size was sufficient                          | Week 2       |
| Odds Ratios               | For interpreting effect magnitude across models                     | Throughout   |

---

### Reflection on Model Fit and Assumptions

We also considered:

- **Model assumptions**, such as the proportional odds assumption in ordinal regression.
- **Potential biases**, including response bias, recall bias, and non-response.
- **Generalizability**, which is quite strong in this dataset due to national sampling.

While we cannot prove causality (cross-sectional data), the strength and direction of our findings strongly align with broader evidence in epidemiology and health equity literature.

---

### Preview of Next Section: Final Discussion

Now that the modeling is complete, we’re ready to move to the **final discussion**. This will include:

- Summary of key findings
- Methodological reflections
- Limitations
- Policy implications
- Personal takeaways

We'll also reflect on how this project helped us **practice real-world statistical reasoning**, and how it connects back to our academic goals in public health and epidemiology.


---

### Final Discussion

This final discussion brings together all the different parts of our analysis — from filtering and cleaning a large public dataset to using statistical models to answer real-world health questions. We reflect not only on our results but also on what this project taught us about data, statistics, and health inequality.

---

### Key Findings

Here’s a brief summary of what we found:

- **Prevalence of Long COVID** is clearly associated with **lower income**, **lower education**, and **barriers to healthcare access**.
- **Severity of Long COVID symptoms** (for those who had it) is also worse among groups with fewer socioeconomic resources.
- **Healthcare access factors** like needing care but not getting it, or delaying care due to cost, had some of the **strongest effects** on symptom severity.
- **Age** and **gender** both influenced outcomes, with older respondents and females slightly more likely to report Long COVID, especially with daily limitations.
- Many of the most meaningful predictors are **structural**, not individual — pointing to systemic inequality as a major component of health outcomes.

Together, these patterns show that Long COVID is not just a medical issue — it’s deeply linked to **social and economic factors**.

---

### Why These Methods?

We chose each method for a specific reason, grounded in the course material:

- We used **logistic regression** to predict the likelihood of Long COVID (binary outcome) and **ordinal logistic regression** to analyze severity (ordered outcome).
- **Chi-square tests** helped us assess relationships between categorical predictors and outcomes — perfect for exploring differences by education level, income category, and care access.
- We used **Cramér’s V** to understand the effect size for associations, as recommended in lectures on categorical data analysis.
- **Mantel–Haenszel tests** allowed us to account for stratification (e.g. age groups), just like we did in lab work involving confounding.
- A **power analysis** reassured us that our sample size (11,000+) was large enough for even small effects to be statistically detectable.
- Odds Ratios helped us **quantify relationships**, which is a key tool for epidemiologists making policy or medical decisions.

These tools weren't just “theoretically appropriate” — they made the results **understandable and useful** in real-world terms.

---

### Limitations

No analysis is perfect, and we were very aware of the following:

- **Self-reported data**: The NHIS survey relies on how people remember and interpret their experiences, which introduces **recall bias** and **response bias**.
- **Cross-sectional design**: Because this dataset captures one point in time, we cannot make strong claims about **causality**.
- **Simplification of categories**: For some variables (like income or education), we collapsed categories or used proxies. This helped statistical power but may lose nuance.
- **Unmeasured confounders**: Factors like vaccination status, comorbidities, mental health, or work exposure weren’t available in our version of the dataset but could strongly affect Long COVID outcomes.

That said, these limitations are **typical in public health research**, and we did our best to mitigate them (e.g. using stratification or interaction terms where appropriate).

---

### Connection to the Course

This project directly used many concepts from our **Biostatistics and Epidemiology module**:

- **Variable filtering and cleaning** (Week 1–2)
- **Descriptive summaries and visualization** (Week 2–3)
- **Chi-squared tests and effect size** (Week 3–4)
- **Regression modeling for binary and ordinal outcomes** (Week 6–7)
- **Multivariable model building and interaction terms** (Week 7–8)
- **Power analysis and interpretation** (Week 2, 9)
- **Understanding structural health disparities** (Guest lectures, discussions)

We got to practice not just applying formulas, but **thinking critically** about what they mean, when to use them, and how to explain them to a general audience.

---

### Personal Reflection

Honestly, this project made statistics feel more **real** to us than any textbook example ever did.

We weren’t just learning R syntax — we were **asking a real question**, and using real data to answer it. That was exciting and also sometimes overwhelming.

One of the most interesting takeaways for us was how **data cleaning and variable selection** are almost more important than the model itself. Getting the structure right — choosing the right categories, encoding missing data correctly — was half the battle.

Another thing we learned was the **power of visualization**. Some relationships only became obvious when we saw them in a plot — especially with severity scores and categorical breakdowns. It reminded us that **good communication** is part of good science.

---

### What We’d Do Next

If we had more time or data, we would:

- Add vaccination status, comorbidities, or employment type as covariates
- Use a **time-to-event model** or panel data, if available, to understand Long COVID progression
- Conduct stratified analyses by **race/ethnicity**, which we left out due to complexity
- Explore **geographic patterns**, if ZIP or region data was available

There’s definitely more to study here — Long COVID is a new and evolving condition, and its social side is just beginning to be understood.

---

### Final Thought

This project made us realize how important it is to think **beyond the disease** — to look at how health is shaped by housing, income, education, and access.

It’s also made us more confident in using statistical tools to approach messy real-world problems. We feel way more ready now to tackle independent research or contribute to policy work that’s grounded in **both data and empathy**.



